Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-10 paper/section-general-framework.tex
Fix typesetting of absolute value operation
Rearrange the construction of non-compositional relational environments a bit to shorten it

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-operations}
  The two-dimensional geometry system has a sort for each of the
  geometric groups mentioned in \autoref{sec:motivating-examples}, so
  $\SortSet = \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$.
  %For the index operations, 
  We have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$, and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
  %  with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
%  The intended interpretations of the top three pairs of operations
%  are group unit, group combination and group negation,
%  respectively. 
%  When we discuss equational theories on index expressions 
%in \autoref{sec:type-equality} 
%we will impose the (abelian) group laws. For this example, 
We also have $\PrimTypeSet = \{ \tyPrimNm{vec},
  \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
           [\SynGL{2}, \SynTransl{2}]$ and
           $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

@@

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

We give two constructions of families of sets of relational
environments that allow us to derive useful consequences of
\thmref{thm:abstraction}.

@@

To satisfy the hypothesis of \thmref{thm:abstraction} we still need to
prove, for each instantiation of our general framework, that the
program-level primitive operations preserve all the relations. This is
the content of \lemref{lem:geom-environments},
\lemref{lem:monoid-ops-related},
\lemref{lem:environments-information-flow} and
\lemref{lem:metric-environments}.

\paragraph{Non-compositional Relational Environments}
We now construct relational environments that can be used to prove
non-definability results. To this end, consider the typing
$t\mathord:\SynTransl{2}; \Gamma_{\Geom} \vdash M : \tyPrim{vec}{t +
  t} \to \tyPrim{vec}{t}$ involving vectors indexed by translations.
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: given the operations on
vectors in \autoref{sec:affine-vector-ops}, there is no way of
cancelling the additional $t$ in the argument type to get the result
type. More generally, the vector operations only allow us to combine
indices according to the group operations, so if the index of the
result type of a function lies outside the free abelian group
generated by the indices of the input types, then we cannot write a
program of this type. 
We now develop this idea to generate a
family of sets of relational environments. The key idea is to assign
different relational interpretations to types with indices that lie
outside some finitely generated collection of index expressions.

These relational environments are non-compositional: elements of
$\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
these relational environments will only be
\emph{pre-substitutive}. Therefore our non-definability results are
subject to the quantifier-free conditions in \thmref{thm:abstraction}.

@@

We construct relational environments parameterised by sets of terms
$S$. We assign the equality relation to primitive types whose index
arguments are generated up to equivalence by $S$. For terms whose
index arguments are not in $\Gen^\equiv(S)$, we require a default
binary relation $R^\bullet_{\tyPrimNm{X}} \subseteq
\Eq_{\tyPrimSem{\tyPrimNm{S}}}$. We define the non-compositional
family of sets of relational environments as
$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
\exists s. \Delta \vdash e : s \}$, where
\begin{displaymath}
  \rho^S\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
    \begin{array}{ll}
      \mathrm{Eq}_{\tyPrimSem{\tyPrimNm{X}}} & \textrm{if all}\ e_i \in \Gen^\equiv(S) \\
      R^\bullet_{\tyPrimNm{X}} & \textrm{otherwise}
    \end{array}
  \right.
\end{displaymath}

>>>>

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-operations}
  The two-dimensional geometry system has a sort for each of the
  geometric groups mentioned in \autoref{sec:motivating-examples}, so
  $\SortSet = \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$.
  %For the index operations, 
  We have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$, and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant, and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, \abs{\cdot} \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, and
  %  with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(\abs{\cdot}) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
%  The intended interpretations of the top three pairs of operations
%  are group unit, group combination and group negation,
%  respectively. 
%  When we discuss equational theories on index expressions 
%in \autoref{sec:type-equality} 
%we will impose the (abelian) group laws. For this example, 
We also have $\PrimTypeSet = \{ \tyPrimNm{vec},
  \tyPrimNm{real} \}$, with $\primTyArity(\tyPrimNm{vec}) =
           [\SynGL{2}, \SynTransl{2}]$ and
           $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

@@

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples} we assumed 
  %that 
  various equational axioms %hold 
  for indexing expressions standing for
  elements of geometric groups. Assuming the abelian group axioms for
  translations we can formalise this in our %general 
  framework:
  %We can now make this assumption formal
  %in our general framework. Semantically, translations form an abelian
  %group under addition, so we assume the abelian group axioms for
  %translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 0 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 0 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $\abs{\cdot}$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $\abs{\det(\iota_OO)} \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

We give two constructions of families of sets of relational
environments that allow us to derive useful consequences of
\thmref{thm:abstraction}. To satisfy the hypothesis of
\thmref{thm:abstraction} we still need to prove, for each
instantiation of our general framework, that the program-level
primitive operations preserve all the relations in our
constructions. This is the content of Lemmas
\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
and \ref{lem:metric-environments}.

@@

\paragraph{Non-compositional Relational Environments}
We now construct relational environments that can be used for
non-definability results. To this end, consider the typing
$t\mathord:\SynTransl{2}; \Gamma_{\Geom} \vdash M : \tyPrim{vec}{t +
  t} \to \tyPrim{vec}{t}$ involving vectors indexed by translations.
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: with the operations on
vectors in \autoref{sec:affine-vector-ops}, there is no way of
cancelling the additional $t$ in the argument type to get the result
type. More generally, the vector operations only allow us to combine
indices according to the group operations, so if the index of the
result type of a function lies outside the free abelian group
generated by the indices of the input types, then we cannot write a
program of this type.  The key idea is to assign different relational
interpretations to types with indices that lie outside some generated
collection of index expressions.

@@

Relational environments are parameterised by sets of terms $S$:
\begin{displaymath}
  \rho^S\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
    \begin{array}{ll}
      \mathrm{Eq}_{\tyPrimSem{\tyPrimNm{X}}} & \textrm{if all}\ e_i \in \Gen^\equiv(S) \\
      R^\bullet_{\tyPrimNm{X}} & \textrm{otherwise}
    \end{array}
  \right.
\end{displaymath}
Thus we assign the equality relation to primitive types whose index
arguments are generated up to equivalence by $S$. For types whose
indices are not in $\Gen^\equiv(S)$, we require a binary relation
$R^\bullet_{\tyPrimNm{X}} \in \Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
$\relEnv{E}_{R^\bullet}(\Delta) = \{ \rho^S \sepbar \forall e \in S.\
\exists s. \Delta \vdash e : s \}$.

We have not been able to prove that this family of sets of relational
interpretations is substitutive. Therefore our non-definability
results are subject to the quantifier-free conditions in
\thmref{thm:abstraction}.

% Hence we must make do with the
% following theorem:

% These relational environments are non-compositional: elements of
% $\relEnv{E}(\Delta,\Delta')$ are not purely constructed from elements
% of $\relEnv{E}(\Delta)$ and $\relEnv{E}(\Delta')$. As a consequence,
% these relational environments will only be
% \emph{pre-substitutive}.
